---
title: "Ross_Iterated_Filtering"
author: "Zhanhao Zhang"
date: "May 9, 2019"
output: html_document
---

## References
[Getting Started With Pomp](https://kingaa.github.io/pomp/vignettes/getting_started.html)

## Required Inputs
data: a csv with one column for time and some other columns for variables.  
hyper-parameters:  
&nbsp;&nbsp;Np: number of particles  
&nbsp;&nbsp;Nmif: number of iterations  
&nbsp;&nbsp;rw.sd: 0 for parameters that are constant, > 0 for parameters we want to fit.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;It also determines the step size of each parameter during the training.

```{r setup, eval=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages Required
```{r, eval=FALSE}
library(pomp2)
library(magrittr)
library(subplex)
library(foreach)
library(doParallel)
library(ggplot2)
library(dplyr)
```

## Demo (Ross-Macdonald Model)
Registers workers, change the 4 to a larger number if you
have a more powerful computer.
```{r,eval=FALSE}
# Registers workers, change the 4 to a larger number if you
# have a more powerful computer.
registerDoParallel(cores=4)
```

Read the data
``` {r, eval=FALSE}
output_read <- read.csv("output1.csv")
```
The column names of data should be different from the statenames
which is required by pomp
```{r, eval=FALSE}
colnames(output_read) <- c("time", "M_c", "Y_c", "Z_c", "I_c")
statenames <- c("M", "Y", "Z", "I", "K", "ER")
paranames <- c("p", "a", "l", "r", "b", "H", "E", "c")
```
parameter values
```{r, eval=FALSE}
params <- c(E=11, r=0.005, b=0.55, a=0.27, c=0.15, H=2000)
```
initial values for each variable  
K and ER are helper variables that are needed in the step function  
```{r, eval=FALSE}
Csnippet("M=468.9; Y=0; Z=0; I=340;K=0; ER=0;") -> init
```
step functions  
the variables will be updated accordingly at each time step  
  
step function -- stochastic version  
using pseudocode  
```{r, eval=FALSE}
Csnippet("
  M += l;
  M *= p;
  Y *= p;
  double Y_0 = a*K*(M-Y);
  Y += Y_0;
  Z = pow(p, E)*Y;
  I = I - r*I + b*ER*(H - I);

  I = I > 0 ? I : 0;
  M = M > 0 ? M : 0;
  Y = Y > 0 ? Y : 0;
  Z = Z > 0 ? Z : 0;

  K = c*I/H;
  ER = a*Z/H;
") -> step
```
step function -- deterministic version  
converted from Ross-Macdonald equation, approximated using taylor expansion
```{r, eval=FALSE}
Csnippet("
  double di1 = r*dt*I;
  double di2 = b*ER*(H-I)*dt;
  double dm1 = p*M*dt;
  double dm2 = -l*(1+p*dt)*dt;
  double dy1 = p*Y*dt;
  double dy2 = a*K*(M-Y)*dt;
  double dz1 = p*Z*dt;
  double dz2 = a*K*(M-Y)*dt;
  
  I += di1 - di2;
  M += dm1 - dm2;
  Y += dy1 - dy2;
  Z += dz1 - dz2;

  I = I > 0 ? I : 0;
  M = M > 0 ? M : 0;
  Y = Y > 0 ? Y : 0;
  Z = Z > 0 ? Z : 0;

  K = c*I/H;
  ER = a*Z/H;
") -> step_det
```
rmeasure, indicates how each variable in the data are generated from the data we are estimating. In this case, each variable we estimate should be the same as the corresponding variable in the original data.
```{r, eval=FALSE}
Csnippet("I_c=I; M_c=M; Y_c = Y; Z_c = Z;") -> rmeas
```
dmeasure, tells the likelihood of each set of estimated values  
Here, I assume they should center around the value in the original data  
The farther they are from the original data, the less likely it is
```{r, eval=FALSE}
Csnippet("
         double di = dnorm(I - I_c, 0, 10, 0);
         double dm = dnorm(M - M_c, 0, 10, 0);
         double dy = dnorm(Y - Y_c, 0, 10, 0);
         double dz = dnorm(Z - Z_c, 0, 10, 0);
         lik = di*dm*dy*dz;") -> dmeas
```
Construct a pomp object 
  
initialize pomp with the Csnippets we defined in the previous part
```{r, eval=FALSE}
output_read %>%
  pomp(
    times="time", t0=0,
    rinit=init, rmeasure=rmeas,
    rprocess=discrete_time(step,delta.t=1),
    statenames=statenames,
    paramnames=paranames
  ) -> output_pomp
```
Code the deterministic skeleton, which is identical to our ODE,
since we don't have random variables directly appears in the ODE
```{r, eval=FALSE}
output_pomp %>%
  pomp(
    skeleton=map(
      Csnippet("DM=l*(1-p)-p*M; DY=a*K*(M-Y)-p*Y;
                DZ=pow(p, E)*DY;DI=b*ER*(H-I)-r*I;"),
      delta.t=1
    ),
    paramnames=paranames,
    statenames=statenames
  ) -> output_pomp
```

Estimate the parameters using the subplex package
Currently I don't use it anymore, but I leave it here for debugging
If this function gives an output without error messages, then we
haven't made any syntactic errors in the previous parts
```{r, eval=FALSE}
output_pomp %>%
  traj_objfun(
    est=c("p", "l"),
    params=c(params,p=1, l=1),
    dmeasure=dmeas, statenames=statenames
  ) -> ofun

subplex(c(1.1, 1.1),fn=ofun) -> fit
fit 
```
Generate nseq number of random guesses of each paramter
with their lower bound and upper bound
For the paramters we passed in as constants (eg. N)
We can just set both of their lower and upper bounds to their values (eg. N = 500)
```{r, eval=FALSE}
sobolDesign(
  lower=c(p=0, l=0, E=11, r=0.005, b=0.55, a=0.27, c=0.15, H=2000),
  upper=c(p=1, l=100, E=11, r=0.005, b=0.55, a=0.27, c=0.15, H=2000),
  nseq=200
) -> guesses
```
Initialize the object for particle filter using random guesses we generated  
Np is the number of particles we use  
Nmif is the number of iterations  
rw.sd is the standard deviation of a paramter when sample it in the particle filter algorithm  
rw.sd of a constant should be set to 0, so that its value won't change
```{r, eval=FALSE}
output_pomp %>%
  mif2(
    params=guesses[1,],
    Np=200,
    Nmif=150,
    dmeasure=dmeas,
    partrans=parameter_trans(log=c("p", "a", "l", "r", "b", "H", "E", "c")),
    cooling.fraction.50=0.5,
    rw.sd=rw.sd(E=0, r=0, b=0, a=0, c=0, H=0, p=0.01, l=0.01),
    paramnames=paranames,
    statenames=statenames,
    filter.traj=TRUE,
    pred.mean=TRUE,
    pred.var=TRUE
  ) -> mf1
```
Do a global search
```{r, eval=FALSE}
foreach(guess=iter(guesses, "row"),
        .combine=c, .packages=c("pomp2"),
        .errorhandling="remove", .inorder=FALSE) %dopar% {
          mf1 %>% mif2(params=guess)
        } ->mifs
```
Get the likelihood of these estimates
```{r, eval=FALSE}
foreach (mf=mifs,
         .combine=rbind, .packages=c("pomp2"), 
         .errorhandling="remove", .inorder=FALSE) %dopar% {
           
           replicate(5, 
                     mf %>% pfilter() %>% logLik()
           ) %>%
             logmeanexp(se=TRUE) -> ll
           
           data.frame(as.list(coef(mf)),loglik=ll[1],loglik.se=ll[2])
           
         } -> estimates
```
More search efforts
```{r, eval=FALSE}
estimates %>% select(-loglik, -loglik.se) -> starts
foreach (start=iter(starts,"row"),
         .combine=rbind, .packages=c("pomp2"),
         .errorhandling="remove", .inorder=FALSE) %dopar% {
           
           mf1 %>%
             mif2(
               params=start,
               partrans=parameter_trans(log=c("p", "a", "l", "r", "b", "H", "E", "c")),
               rw.sd=rw.sd(E=0, r=0, b=0, a=0, c=0, H=0, p=0.01, l=0.01),
               paramnames=paranames
             ) %>%
             mif2() -> mf
           
           replicate(5, 
                     mf %>% pfilter() %>% logLik()
           ) %>%
             logmeanexp(se=TRUE) -> ll
           
           data.frame(as.list(coef(mf)),loglik=ll[1],loglik.se=ll[2])
         } -> r_prof
```
Get the MLE and draw the simulation plots
```{r, eval=FALSE}
r_prof[which.max(r_prof[,9]),][1:8] -> mle

mlepomp <- as(mifs[[1]], 'pomp')
coef(mlepomp) <- mle

mlepomp2 <- as(mifs2[[1]], 'pomp')
coef(mlepomp2) <- mle

# Simulation plots for I
mlepomp %>%
  simulate(nsim=1,format="data.frame",include.data=TRUE) %>%
  ggplot(mapping=aes(x=time,y=I_c,group=.id,alpha=(.id=="data")))+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.2),
                     labels=c(`FALSE`="simulation",`TRUE`="data"))+
  labs(alpha="")+
  geom_line()+
  theme_bw()
ggsave("plots/sim_I_dt=1_Nmif=150_Np=200.png")
# Simulation plots for M
mlepomp %>%
  simulate(nsim=3,format="data.frame",include.data=TRUE) %>%
  ggplot(mapping=aes(x=time,y=M_c,group=.id,alpha=(.id=="data")))+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.2),
                     labels=c(`FALSE`="simulation",`TRUE`="data"))+
  labs(alpha="")+
  geom_line()+
  theme_bw()
ggsave("plots/sim_M_dt=1_Nmif=150_Np=200.png")
# Simulation plots for Y
mlepomp %>%
  simulate(nsim=3,format="data.frame",include.data=TRUE) %>%
  ggplot(mapping=aes(x=time,y=Y_c,group=.id,alpha=(.id=="data")))+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.2),
                     labels=c(`FALSE`="simulation",`TRUE`="data"))+
  labs(alpha="")+
  geom_line()+
  theme_bw()
ggsave("plots/sim_Y_dt=1_Nmif=150_Np=200.png")
# Simulation plots for Z
mlepomp %>%
  simulate(nsim=3,format="data.frame",include.data=TRUE) %>%
  ggplot(mapping=aes(x=time,y=Z_c,group=.id,alpha=(.id=="data")))+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.2),
                     labels=c(`FALSE`="simulation",`TRUE`="data"))+
  labs(alpha="")+
  geom_line()+
  theme_bw()
ggsave("plots/sim_Z_dt=1_Nmif=150_Np=200.png")
```